``Interestingness'' in Mathematical Theory Exploration

### 2018-10-15 PhDThesis/background_haskell.tex
More content

We mostly focus our attention on the Haskell programming language, both as an
implementation vehicle and as our representation of functions, properties, etc.
to explore. We choose Haskell since it combines formal, logical underpinnings
which aid reasoning (compared to more popular languages like Java or C), yet it
is still popular enough to sustain a rich ecosystem of tooling and a large
corpus of existing code (compared to more formal languages like Coq or
Isabelle).

@@

%% FIXME: MERGE THE FOLLOWING WITH THE ABOVE

@@

We decided to focus on theory exploration in the Haskell programming language as
it has mature, state-of-the-art implementations (\qspec{} \cite{QuickSpec} and
\hspec{} \cite{claessen2013automating}). This is evident from the fact that the
state-of-the-art equivalent for Isabelle/HOL, the \textsc{Hipster}
\cite{Hipster} system, is actually implemented by translating to Haskell and
invoking \hspec{}.

Haskell is well-suited to programming language research; indeed, this was a goal
of the language's creators \cite{marlow2010haskell}. Like most
\emph{functional} programming languages, Haskell builds upon $\lambda$-calculus,
with extra features such as a strong type system and ``syntactic sugar'' to
improve readability. For simplicity, we will focus on an intermediate
representation of the \textsc{GHC} compiler, known as \emph{GHC Core}, rather
than the relatively large and complex syntax of Haskell proper. Core is based on
\fc{}, but for our machine learning purposes we are mostly interested in its
syntax; for a more thorough treatment of \fc{} and its use in GHC, see
\cite[Appendix C]{sulzmann2007system}.

The sub-set of Core we consider is shown in Figure \ref{fig:coresyntax};
compared to the full language \footnote{As of GHC version 7.10.2, the latest at
  the time of writing.} our major restriction is to ignore type hints (such as
explicit casts, and differences between types/kinds/coercions). For brevity, we
also omit several other forms of literal (machine words of various sizes,
individual characters, etc.), as their treatment is similar to those of strings
and numerals. We will use quoted strings to denote names and literals,
e.g. \texttt{Local "foo"}, \texttt{Global "bar"}, \texttt{Constructor "Baz"},
\texttt{LitStr "quux"} and \texttt{LitNum "42"}, and require only that they can
be compared for equality.

@@

Figure \ref{fig:haskellexample} shows some simple Haskell function definitions,
along with a custom datatype for Peano numerals. The translation to our Core
syntax is routine, and shown in Figure \ref{fig:coreexample}. Although the Core
is more verbose, we can see that similar structure in the Haskell definitions
gives rise to similar structure in the Core; for example, the definitions of
\hs{odd} and \hs{even} are identical in both languages, except for the
particular identifiers used. It is this close correspondence which allows us to
analyse Core expressions in place of their more complicated Haskell source.

>>>>

We mostly focus our attention on the Haskell programming language, both as an
implementation vehicle and as our representation of functions, properties, etc.
to explore. We choose Haskell since it combines formal, logical underpinnings
which aid reasoning (compared to more popular languages like Java or C), yet it
is still popular enough to sustain a rich ecosystem of tooling and a large
corpus of existing code (compared to more formal languages like Coq or
Isabelle). Haskell is well-suited to programming language research; indeed, this
was a goal of the language's creators \cite{marlow2010haskell}. Like most
\emph{functional} programming languages, Haskell builds upon $\lambda$-calculus,
with extra features such as a strong type system and ``syntactic sugar'' to
improve readability.

@@

@@

Further evidence of Haskell's suitability for theory exploration is given by the
fact that the state-of-the-art implementation for Isabelle/HOL, the
\textsc{Hipster}~\cite{Hipster} system, is actually implemented by translating
to Haskell and invoking \hspec{}~\cite{claessen2013automating}.

Whilst we will use normal Haskell code in this text and many of our examples,
for simplicity we perform many of our analyses on an intermediate representation
of the \textsc{GHC} compiler, known as \emph{GHC Core}, rather than the
relatively large and complex syntax of Haskell proper. Core is based on \fc{},
which is described in detail in~\cite[Appendix C]{sulzmann2007system}.

Core contains explicit type annotations and coercions, which we can omit as they
have no effect on runtime behaviour. The resulting sub-set of Core\footnote{As
  of GHC version 7.10.2, the latest at the time of writing.} is shown in
Figure~\ref{fig:coresyntax}; for brevity, we also omit several other forms of
literal (machine words of various sizes, individual characters, etc.), since
their treatment is similar to those of strings and numerals. We use quoted
strings to denote names and literals, e.g. \texttt{Local "foo"},
\texttt{Global "bar"}, \texttt{Constructor "Baz"}, \texttt{LitStr "quux"} and
\texttt{LitNum "42"}, and require only that they can be compared for equality.

@@

Figure~\ref{fig:haskellexample} shows some simple Haskell function definitions,
along with a custom datatype for Peano numerals. The translation to Core syntax
is routine, and shown in Figure~\ref{fig:coreexample}. Although the Core is more
verbose, we can see that similar structure in the Haskell definitions gives rise
to similar structure in the Core; for example, the definitions of \hs{odd} and
\hs{even} are identical in both languages, except for the particular identifiers
used. It is this close correspondence which allows us to analyse Core
expressions in place of their more complicated Haskell source.

### 2018-10-15 PhDThesis/background_theoryexploration.tex
More content

We focus on \qspec{} \cite{QuickSpec}, which conjectures equations about
Haskell code (these may be fed into another tool, such as \hspec{}, for
proving). These conjectures are arrived at through the following stages:

>>>>

We focus on \qspec{}~\cite{QuickSpec}, which conjectures equations about
Haskell code (these may be fed into another tool, such as \hspec{}, for
proving). These conjectures are arrived at through the following stages:

### 2018-10-15 PhDThesis/litreview_premiseselection.tex
More content

% TODO
\cite{kuhlwein2012overview}

One fruitful application of machine learning methods to formal mathematics is
\emph{relevance filtering}, which is used for \emph{premise selection} in
automated theorem provers.

@@

The Multi-output ranking (\textsc{MOR}) implementation uses a support vector
machine (SVM) to select relevant axioms from the Mizar Mathematical Library for
use by the Vampire ATP system~\cite{alama2014premise}.
%TODO: describe the kernel, as that's the interesting bit.

>>>>

One fruitful application of machine learning methods to formal mathematics is
\emph{relevance filtering}, which is used for \emph{premise selection} in
automated theorem provers~\cite{kuhlwein2012overview}.

@@

Premise selection has also been applied to the re-proving problem in other
systems, such as the Multi-Output Ranking (\textsc{MOR}) tool, which uses a
support vector machine (SVM) to select relevant axioms from the Mizar
Mathematical Library for use by the Vampire ATP system~\cite{alama2014premise}.
%TODO: describe the kernel, as that's the interesting bit.

### 2018-10-15 PhDThesis/litreview_staticanalysis.tex
More content

The other category is that of ``standalone'' tools, which are run separately to
building or executing the target code, and whose results do not contribute
directly to the code's output. These include tools for testing, profiling,
verification, standards compliance, etc. Our application of theory exploration
to software libraries fits into this latter category, whose constraints are not
as stringent as those applying to tools like compilers. The only real
restrictions on these tools are that their resource requirements are acceptable
to users, and that output is sufficiently high quality to be worth waiting for
and sifting through.

Some of the simplest code improvement tools are known as \emph{linters}, after
the \textsc{Lint} tool for the C programming language~\cite{Johnson78lint}.
Linters are characterised by parsing the given code and looking for occurrences
of certain syntax patterns, for example definitions which are never used, or
``no-op'' constructs (e.g. \texttt{(x < 0)} when \texttt{x} is of
unsigned/non-negative type, making this comparison always false).

A popular linter for Haskell code is \textsc{HLint}~\cite{mitchell2014hlint},
which focuses mostly on simplifying code; for example by spotting use-cases for
common library functions like \texttt{null x} instead of \texttt{length x == 0}.
Linting can be seen as the inverse of theory exploration: rather than inferring
novel, general relationships from patterns found in the code, linting begins
with known relationships and looks through the code for specific instances.
Whilst linting is a useful programming aid, and fast enough to apply regularly
(e.g. as a background process in a code editor), it is not an open-ended
process: once the predetermined patterns are exhausted, no more improvements
will be found.

At the other end of the code improvement spectrum we find
\emph{superoptimisers}~\cite{}, which can take an
arbitrarily long time to find improvements, but whose deductive power is only
limited by the undecidability of the source language. Superoptimisers, like ,
treat the given code as a specification, and perform an open-ended search
through the space of all programs to find the optimal implementation according
to some quality measure or cost model (often a branch- and effect-free subset of
machine code is used, and measured by counting instructions or
cycles~\cite{massalin1987superoptimizer}). Superoptimisers need some method to
check whether a synthesised program implements the required specification, such
as a test suite and/or a theorem prover.

Proving equivalence of machine code programs is helped by the uniform, finite
nature of their domain (e.g. a fixed set of instructions, operating on a fixed
set of 32bit registers). Higher level languages, like Haskell, are more
difficult to reason about and measure, since their behaviour and efficiency may
depend on unknown runtime parameters. For example, a higher-order function may
choose to memoise its argument, which avoids repeated function calls but uses
more memory and comparison checks; whether this is an optimisation or
pessimisation depends on the runtime data at each call site, including the
function argument (whether it's cheaper than performing a comparison) and the
data it will be applied to (whether it actually contains any repeated values).
In addition to this uncertainty, each function's input domain may be infinite
(such as the natural numbers), and programs may use abstract interfaces and
datatypes with a variety of possible implementations (such as lists or key/value
mappings). All these factors makes effective testing more difficult, and proofs
potentially undecidable.

We can reduce the amount of unknown parameters in a concrete program by
propagating statically-known values, for example using
\emph{supercompilation}~\cite{Turchin:1986:CS:5956.5957}; however this can
dramatically increase code size, as each function may be duplicated into
multiple specialised instances. Efficiency can be calculated
using techniques like cost semantics~\cite{danner2015denotational}, with
constant factors determined through empirical testing. This cannot solve the
problem of optimality depending on the characteristics of runtime data, but it
can expose these dependencies as parameters allowing the user to select
between pareto-optimal strategies.

>>>>

The other category is that of ``standalone'' tools, which are run separately to
building or executing the target code, and whose results do not contribute
directly to the code's output. These include tools for testing, profiling,
verification, standards compliance, etc. The application of theory exploration
to software libraries fits into this latter category, whose constraints are not
as stringent as those applying to tools like compilers. The only real
restrictions on these tools are that their resource requirements are acceptable
to users, and that output is sufficiently high quality to be worth waiting for
and sifting through.

Some of the simplest code improvement tools are known as \emph{linters}, after
the \textsc{Lint} tool for the C programming language~\cite{Johnson78lint}.
Linters are characterised by parsing the given code and looking for occurrences
of certain syntax patterns, for example definitions which are never used, or
``no-op'' constructs (e.g. \texttt{(x~<~0)} when \texttt{x} is of
unsigned/non-negative type, making this comparison always false).

A popular linter for Haskell is \textsc{HLint}~\cite{mitchell2014hlint}, which
focuses mostly on simplifying code; for example by spotting use-cases for common
library functions like \texttt{null x} instead of \texttt{length x == 0}.
Linting can be seen as the inverse of theory exploration: rather than inferring
novel, general relationships from patterns found in the code, linting begins
with known relationships and looks through the code for specific instances.
Whilst linting is a useful programming aid, and fast enough to apply regularly
(e.g. as a background process in a code editor), it is not an open-ended
process: once the predetermined patterns are exhausted, no more improvements
will be found.

At the other end of the code improvement spectrum we find
\emph{superoptimisers}~\cite{massalin1987superoptimizer}, which can take an
arbitrarily long time to find improvements, but whose deductive power is only
limited by the undecidability of the source language. Superoptimisers, such as
\textsc{Souper}~\cite{sasnauskas2017souper} and \textsc{Aha!} (``A Hacker's
Assistant'', developed to accompany \emph{Hacker's
  Delight}~\cite{warren2013hacker}) treat the given code as a specification, and
perform an open-ended search through the space of all programs to find the
optimal implementation according to some quality measure or cost model (often a
branch- and effect-free subset of machine code is used, and measured by counting
instructions or cycles).  Superoptimisers need some method to check whether a
synthesised program implements the required specification, such as a test suite
and/or a theorem prover.

Proving correctness of machine code programs is helped by the uniform, finite
nature of their domain (e.g. a fixed set of instructions, operating on a fixed
set of fixed-size registers). Higher level languages like Haskell are more
difficult to reason about and measure, since their behaviour and efficiency may
depend on unknown runtime parameters. For example, a higher-order function $f$
accepting another function $g$ as an argument may choose to memoise calls to $g$
in order to avoid evaluating identical calls multiple times; however, storing
previous results requires more memory, and each input must be looked up in the
memo table to see if it's been used before. Whether this is an optimisation or
pessimisation depends on the runtime data at each call site, including the
function argument (whether it's cheaper than performing a comparison) and the
data it will be applied to (whether it actually contains any repeated values).
In addition to this uncertainty, each function's input domain may be infinite
(such as the natural numbers), and programs may use abstract interfaces and
datatypes with a variety of possible implementations (such as lists or key/value
mappings). All these factors makes effective testing more difficult, and proofs
potentially undecidable.

We can reduce the amount of unknown parameters in a concrete program by
propagating statically-known values, for example using
\emph{supercompilation}~\cite{Turchin:1986:CS:5956.5957}; however this can
dramatically increase code size, as each function may be duplicated into
multiple specialised instances. Related techniques, like \emph{symbolic
  execution} (a form of \emph{abstract interpretation}), allow approximate
execution of code, where unknown runtime data is represented as opaque
metavariables. Property- and model-checkers are similar, except they work
empirically by performing multiple runs, instantiating metavariables with
synthetic data (\emph{QuickCheck} does this randomly, \emph{SmallCheck}
enumerates values, etc.).

Efficiency can be calculated using techniques like cost
semantics~\cite{danner2015denotational}, with constant factors determined
through empirical testing. This cannot solve the problem of optimality depending
on the characteristics of runtime data, but it can expose these dependencies as
parameters allowing the user to select between pareto-optimal strategies.

### 2018-10-15 PhDThesis/litreview_testing.tex
More content

The \textsc{LazySmallCheck} framework is similar in spirit to
\textsc{SmallCheck}, but utilises Haskell's lazy evaluation to avoid generating
values up front~\cite{reich2013advances}. Instead, a partial value is generated
which will trigger an exception if evaluated. Since catching a Haskell exception
is an \texttt{IO} action, pure properties which inspect such values will always
be aborted, handing control back to \textsc{LazySmallCheck} which checks the
same property on refined versions of the previous value. For example, a lazily
generated \texttt{Tree} may initially be equivalent to \texttt{undefined}. If
the property aborts, it will be run again with inputs equivalent to
\texttt{Leaf}
and \texttt{Node undefined undefined}; if the latter aborts due to inpuecting
the left sub-tree, then \texttt{Node Leaf undefined} and
\texttt{Node (Node undefined undefined) undefined} will be tried (and likewise
for the right sub-tree). This refinement continues until the property no longer
aborts or until a predefined depth limit is reached. If the property returns a
true value when given such partially-defined input then it must succeed for
\emph{all} refinements of that input (since the output was determined without
evaluating that sub-term). If the property returns a false value, the input is a
\emph{minimal} counterexample, since it only specifies those sub-terms which are
required to violate the property; any undefined subterms are irrelevant to the
failure.

\textsc{LazySmallCheck}'s automatic elaboration of free variables is similar in
operation to the use of search procedures in logic programming languages like
Kanren, Prolog (which has its own \textsc{QuickCheck} port) and Curry (a
functional logic language, whose \textsc{EasyCheck} framework explicitly
unifies the concepts of generators and logic
variables~\cite{christiansen2008easycheck}). The use of uninterpreted symbols
is also reminiscent of \emph{symbolic execution}~\cite{}, a form of
\emph{abstract interpretation} where unknown runtime values are represented by
opaque symbols, computation proceeds to arrive at a result, which may be a
complicated expression involving many symbols. Importantly, we can  the in terms of the until the result is
by accumulating
combinations of these symbols. For example, we may not know the output of a function

>>>>

The \textsc{LazySmallCheck} framework is similar in spirit to
\textsc{SmallCheck}, but utilises Haskell's lazy evaluation to avoid generating
values up front~\cite{reich2013advances}. Instead, a partial value is generated
which will trigger an exception if evaluated. Since catching a Haskell exception
is an \texttt{IO} action, pure properties which inspect such values will always
be aborted, handing control back to \textsc{LazySmallCheck} which checks the
same property on refined versions of the previous value. For example, a lazily
generated \texttt{Tree} may initially be equivalent to $\bot$. If the property
aborts, it will be run again with each possible refinement of $\bot$: in this
case inputs equivalent to \texttt{Leaf} and \texttt{Node~$\bot_1$~$\bot_2$}. If
the latter aborts due to inspecting $\bot_1$, then $\bot_1$ will be refined to
give inputs \texttt{Node~Leaf~$\bot_2$} and
\texttt{Node~(Node~$\bot_3$~$\bot_4$)~$\bot_2$} (and likewise if $\bot_2$ had
caused the exception). This refinement continues until the property no longer
aborts or until a predefined depth limit is reached. If the property returns a
true value when given such partially-defined input then it must succeed for
\emph{all} refinements of that input (since the output was determined without
evaluating the undefined sub-terms). If the property returns a false value, the
input is a \emph{minimal} counterexample, since it only specifies those
sub-terms which are required to violate the property; any undefined sub-terms
are irrelevant to the failure.

\textsc{LazySmallCheck}'s automatic elaboration of free variables is similar in
operation to the use of search procedures in logic programming languages like
Kanren, Prolog (which has its own \textsc{QuickCheck} port) and Curry (a
functional logic language, whose \textsc{EasyCheck} framework explicitly
unifies the concepts of generators and logic
variables~\cite{christiansen2008easycheck}). The use of uninterpreted symbols
is also reminiscent of \emph{symbolic execution}~\cite{king1976symbolic}, a form
of \emph{abstract interpretation} where unknown runtime values are represented
by opaque symbols and computation proceeds to arrive at a result, which may be a
complicated expression involving many symbols.
